---
author: "Climate Change Response Program"
date: "2025-03-11"
output: word_document
---

```{r setup, include=FALSE, echo=FALSE}
library(dplyr); library(tidyr);library(knitr)
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)

# Set park variables
SiteID <- "SHEN"
LongName <- "Shenandoah National Park"
CF_selected <- "WarmWet_HotDry"
# CF_selected <- "WarmDry_HotWet"

region.name <- "Northeast"

# Set directories
OutDir <- "C:/Users/arunyon/DOI/CCRP COLLABORATE! - CCRP COLLABORATE!/01 PROJECT Collaboration/Science, Adaptation, Planning/Resource Stewardship Strategies (RSS)/Current Park Folders/2024_SHEN/"

Park_CF_Dir <- "C:/Users/arunyon/DOI/CCRP COLLABORATE! - CCRP COLLABORATE!/01 PROJECT Collaboration/Science, Adaptation, Planning/Resource Stewardship Strategies (RSS)/Current Park Folders/2024_SHEN/SHEN - Climate Futures/"

HistDir <- "C:/Users/arunyon/DOI/CCRP COLLABORATE! - CCRP COLLABORATE!/01 PROJECT Collaboration/Science, Adaptation, Planning/Resource Stewardship Strategies (RSS)/Current Park Folders/2024_SHEN/SHEN - Historical/"


# #Read in RData file
load(paste0(Park_CF_Dir, "input-data/Final_Environment.RData"))

#Read in Historical file
Historical_regression <- read.csv(paste0(HistDir, SiteID,"-Regression Table.csv"))

if(CF_selected == "WarmWet_HotDry") {
  CFs = c("Warm Wet", "Hot Dry") # Pick pair of climate futures.
  CF_abbreviation <- "WW-HD"
  CFDir = paste0(Park_CF_Dir,"WarmWet_HotDry/") # for .csv's
  col <- c("#2B83BA","#D7191C") 
  color.names <- c("blue","red")
} else{
  CFs = c("Warm Dry", "Hot Wet")  # Pick pair of climate futures.
  CF_abbreviation <- "WD-HW"
  CFDir = paste0(Park_CF_Dir,"WarmDry_HotWet/") # for .csv's
  col <- c("#FDAE61","#ABDDA4")
  color.names <- c("orange","green")
}
TableDir = paste0(CFDir,"tables/") # for .csv's
FigDir = paste0(CFDir,"figures/") # for .csv's

D_Annual <- openxlsx::read.xlsx(xlsxFile=paste0(TableDir,SiteID,"_",CF_abbreviation,"_Plot_data.xlsx"),sheet="D_Annual")
AnnualWB <- read.csv(paste0(TableDir,"WB-Annual.csv")) |> 
  left_join(WB_GCMs,by="GCM") |> 
  mutate(sum_d.in = sum_d.mm/ 25.4,
         sum_aet.in = sum_aet.mm / 25.4)
AnnualWB <- AnnualWB |> 
  replace_na(list(CF = "Historical")) |> 
  mutate(CF = factor(CF, levels=c("Historical",CFs)))
Drought.char <- read.csv(paste0(TableDir, "Drought_characteristics.csv"))

#Precip quad direction - saves in-line code
CF1_precip_direction <- if(D_Annual$PrcpIn[2] > 0) {"pos"} else {"neg"}
CF2_precip_direction <- if(D_Annual$PrcpIn[3] > 0) {"pos"} else {"neg"}

#Monthly trends
CF1_precip_inc_direction <- Season_delta |> filter(CF == CFs[1],
                                                   PrcpIn > 0) |> 
  dplyr::select(season)  
CF1_precip_inc_direction <- as.character(CF1_precip_inc_direction$season)
CF1_precip_inc_direction <- if(length(CF1_precip_inc_direction)==4) {"all"} else{CF1_precip_inc_direction}

CF2_precip_inc_direction <- Season_delta |> filter(CF == CFs[2],
                                                   PrcpIn > 0) |> 
  dplyr::select(season)

CF2_precip_inc_direction <- as.character(CF2_precip_inc_direction$season)
CF2_precip_inc_direction <- if(length(CF2_precip_inc_direction)==4) {"all"} else{CF2_precip_inc_direction}



#Translation text

# Historical -- look at state climate summary and fill in reference.
State_summary_statement <- "Temperatures in Virginia have risen more than 1.5 °F since the beginning of the 20th century and total annual precipitation shows a small upward trend, with multiyear values mostly above average since 1995 (Runkle et al. 2022)."
State_summary_reference <- "Runkle, J., K.E. Kunkel, L.E. Stevens, S.M. Champion, B.C. Stewart, R. Frankson, W. Sweet, and S. Rayne, 2022: Virginia State Climate Summary 2022. NOAA Technical Report NESDIS 150-VA. NOAA/NESDIS, Silver Spring, MD, 5 pp."

```

---
title: `r LongName` RSS Climate Summary
---

***This template has all of the park climate information filled in. The project lead and regional climate staff need to fill in the park-specific sensitivities section based on concerns raised during the RSS process.***

## Historical and future climate

`r LongName` (`r SiteID`) has already experienced significant and rapid human-caused climate change. (USGCRP 2023). `r State_summary_statement` Both maximum and minimum annual temperatures in `r SiteID` have significantly increased since 1970 and overnight temperatures (daily minimum) have increased at a rate of `r round(Historical_regression[6,3]/10,1)` °F per decade since 1970. Average annual precipitation has `r if(Historical_regression[3,3] > 0) {print("increased")} else {print("decreased")}` `r round(Historical_regression[3,3]/10,1)` inches per decade over the 20th century until present and there have been critical shifts in the style of the precipitation events. Extreme rainfall events have increased regionally and can change the frequency of flood peaks on tributary systems of `r SiteID`, with implications for ***[REGIONAL COORDINATOR FILL POTENTIAL PARK-SPECIFIC IMPLICATIONS, E.G., geomorphology, and sediment transport]***. These sorts of climate changes frequently act as stress multipliers and exacerbate existing threats to natural resources, historic structures, and cultural landscapes.

All climate models project warming temperatures at `r SiteID` by ca. 2050 relative to a 1979-2012 baseline (Analyses: NPS Climate Change Response Program [CCRP]; Figure 1). Future projections for precipitation are mixed; most models project increasing average annual precipitation relative to the historical period while some project decreasing precipitation. CCRP produced two divergent and plausible climate futures specific to `r SiteID` for consideration during the RSS process. These include a "`r CFs[1]`" and "`r CFs[2]`" climate future. Under both futures temperatures increase but differ in the magnitude of warming. The futures also differ in whether they project increasing or decreasing annual precipitation. Considering different scenarios of climate change at `r SiteID` is intended to allow planning for each scenario, accounting for uncertainty in the future conditions the park is likely to experience in management plans (Lawrence et al. 2021).

Specifically, under a "`r CFs[1]`" climate future, mean annual temperatures are projected to rise `r round(D_Annual$TavgF[2],1)` °F by 2050, while mean annual precipitation `r ifelse(CF1_precip_direction=="pos","increases","decreases")` by `r round(D_Annual$PrcpIn[2],1)`” (`r round(D_Annual$PrcpIn[2]/D_Annual$PrcpIn[1]*100,1)`% from baseline). Under a "`r CFs[2]`" climate future, mean annual temperatures are projected to increase `r round(D_Annual$TavgF[3],2)` °F (Figure 1) and mean annual precipitation `r ifelse(CF2_precip_direction=="pos","increases","decreases")` by `r round(D_Annual$PrcpIn[3],1)`” (`r round(D_Annual$PrcpIn[3]/D_Annual$PrcpIn[1]*100,1)`%). Both climate futures project a decrease in days below freezing (`r round(D_Annual$UnderColdTemp[2] / D_Annual$UnderColdTemp[1] * 100, 1)`% to `r round(D_Annual$UnderColdTemp[3] / D_Annual$UnderColdTemp[1] * 100, 1)`%) and an increase in days above the 95th percentile for heat (`r round(HistTmaxHigh, 2)` °F). Such changes in temperature will affect species ranges. They will exclude some species dependent on cooler conditions while allowing further range expansions of sub-tropical species, including non-native exotics.

***RA'S THE FOLLOWING TEXT SHOULD BE MODIFIED AND MERGED INTO ONE PARAGRAPH***

Models also project precipitation may occur during higher-intensity one or two-day events throughout the `r region.name`, due to the higher water vapor content of the atmosphere with increasing temperature (USGCRP 2023).

*RA's to insert seasonal text and combine with previous paragraph. Example below.*

`r paste0(CFs[1], " precipitation is increasing in ", CF1_precip_inc_direction, " seasons.")` `r paste0(CFs[2], " precipitation is increasing in ", CF2_precip_inc_direction, " seasons.")`

Seasonal precipitation is projected to increase in the "`r CFs[1]`" climate future across all seasons. Whereas, seasonal precipitation is projected to decrease across all seasons except spring for the "`r CFs[2]`" climate future.

Overall, climate change is anticipated to intensify the hydrologic cycle at `r SiteID`, increasing the intensity and frequency of extreme events including drought and deluge.

*RA's to modify drought text using the table below and combine with previous paragraph. Delete drought table. Example below.*

```{r echo=FALSE}
knitr::kable(Drought.char, format="markdown")
```

For example, drought intensity and severity are projected to increase under the "`r CFs[2]`" climate future where short but intense “flash droughts” become more common. Drought conditions are not as severe in the "`r CFs[1]`" climate future when extreme precipitation is a greater concern. However, even in this wet scenario, drought conditions remain a concern because even if annual precipitation remained the same, higher temperatures increase evapotranspiration rates, decreasing soil moisture during dry periods.

***The project lead and regional climate staff fill in the park-specific sensitivities section based on concerns raised during the RSS process, here. [Examples can be found here.](https://doimspp.sharepoint.com/sites/nps-waso-ccrp/Shared%20Documents/Forms/AllItems.aspx?id=%2Fsites%2Fnps%2Dwaso%2Dccrp%2FShared%20Documents%2F01%20PROJECT%20Collaboration%2FScience%2C%20Adaptation%2C%20Planning%2FResource%20Stewardship%20Strategies%20%28RSS%29%2FExample%20Climate%20Summaries&viewid=a459d1c3%2Dd72a%2D4fec%2D9adc%2D0cc2a4094eba)***

```{r echo=FALSE}
fig <- paste0(FigDir, "TavgF-PrcpIn-Timeseries.png")

```

![**Figure 1.** Historical (gray) and future projections (`r color.names[1]` and `r color.names[2]` lines) for annual mean temperature and precipitation in `r SiteID`. The lines running through the time series is the 10-year running average. Gray shaded area represents the future time period used to average future changes from the historical baseline. See Runyon et al. (2024) for details.](%60r%20fig%60)

## References

`r State_summary_reference`

Lawrence, D. J., A. N. Runyon, J. E. Gross, G. W. Schuurman, and B. W. Miller. 2021. Divergent, plausible, and relevant climate futures for near- and long-term resource planning. Climatic Change 167:38.

Runyon, A.N., J.E. Gross, G.W. Schuurman, D.J. Lawrence, and J.H. Reynolds. 2024. Methods for assessing climate change exposure for national park planning. Natural Resource Report No. NPS/NRSS/CCRP/NRR—2024/XXXX). National Park Service.

USGCRP, 2023: Fifth National Climate Assessment. Crimmins, A.R., C.W. Avery, D.R. Easterling, K.E. Kunkel, B.C. Stewart, and T.K. Maycock, Eds. U.S. Global Change Research Program, Washington, DC, USA. <https://doi.org/10.7930/NCA5.2023>
